Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-23 second-draft/pnas-word-order-si.tex
SI

@@

In Figure~\ref{fig:pareto-per-lang}, we show the predictability-parseability planes for every one the 51 languages, together with Pareto frontiers estimated from optimized grammars.\footnote{Note that in a few languages, the real grammar is at a position slightly \emph{beyond} the estimated Pareto frontier. Due to use of stochastic gradient descent in optimization, there is some amount of noise in the optimization process.}

We z-transformed on the level of individual languages, normalizing the mean and SD parseability and predictability of the (1) real grammar, (2) the mean of predictability and parseability of all random grammars, (3) the grammar optimized for efficiency (at $\lambda =0.9$, see Section~\ref{sec:lambda}), (4) grammar optimized for parseability only, and (5) grammar optimized for predictability only.

Within each language, we estimate the Pareto curve based on the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section~\ref{sec:lambda}), (2) parseability, and (3) predictability.

Figure (REF) in the main paper shows the average of these per-language plots.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars. All data are $z$-scored.}\label{fig:pareto-per-lang}
\end{figure}

In the main paper, we tested whether real grammars are more efficient than the mean of random grammars, using a $t$-test.
We also conducted the analysis using a Binomial test (one-sided), testing whether the real gramar is more efficient than the \emph{median} of random grammars, avoiding any distributional assumption on the random grammars.
As before, we used Hochberg's step-up procedure (Note that the tests for different languages are independent, as different random grammars are evaluated for each language), with $\alpha=0.05$.
In this analysis, real grammars improved in parseability for 80\% of languages, in predictability for 69\% of languages, and in either of both in 92\% of languages ($p<0.05$, with Bonferroni correction).
In Table~\ref{fig:pareto-per-lang-stats}, we provide per-language results for the $t$-tests and binomial tests.

@@

In Figure~\ref{fig:corr-eff-corr}, we plot efficiency, parseability, and predictability (parseability and predictability $z$-scored within language) as a function of the number of satisfied correlations, for the real grammars of the 51 languages.

@@

We show predictions for the eight correlations on the level of individual languages in Figure~\ref{fig:per-lang}.

We obtained predictions for individual languages and each of the eight relations as follows.
For each language and each of the objective functions (efficiency, predictability, parseability), we considered the optimized grammar that yielded the best value of this objective function among the eight optimized grammars (i.e., the grammar where the optimization procedure had been most successful).
We interpreted this grammar as verb-object or object-verb depending on the order in the real grammar of the language.

@@

Note that the Bayesian analysis  also estimates a posterior distribution of the number of satisfied correlations (see Figure~\ref{fig:posterior}), providing an elegant solution to the multiple-comparisons problem arising from analysing the eight correlation.

@@

In Figure~\ref{fig:posterior}, we plot the posterior number of correlations predicted to hold in most optimized grammars, as obtained from the Bayesian regression.
In addition to grammars optimized for efficiency, we also report the result for grammars optimized for predictability and for parseability alone.
Efficiency predicts all eight correlations with high posterior probability; predictability and parseability alone do not.

\begin{figure}[ht]
	\begin{center}
	\includegraphics[width=0.98\textwidth]{../results/correlations/figures/posterior-satisfied-universals-together-large-three.pdf}
	\end{center}
	\caption{Posterior of the number of correlations correctly predicted by efficiency and its components, in a Bayesian multivariate mixed-effects logistic regression with random effects for languages and language families.}\label{fig:posterior}
\end{figure}

@@

We considered all UD syntactic relations occurring in at least two of the 51 languages.
In Table~\ref{tab:all-predictions-1}, we present the data for the eight correlations discussed in the main paper, and for those other relations for which the typological literature provides data.
Additionally, in Table~\ref{tab:all-predictions-1} we present data for the other UD relations, for which either no typological data is available, or which are not linguistically meaningful.

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM} for discussion).
This has been suggested as a property of efficient word order in prior work, and, indeed, we found that DLM is predicted by efficiency optimization (Section~\ref{sec:DLM}).
Results in Table~\ref{tab:all-predictions-1} also show that optimizing for DLM makes predictions similar to efficiency optimization. See Section~\ref{sec:DLM} for further discussio for further discussionn.

@@

\footnotetext{The \emph{aux} syntactic relation in UD has the auxiliary (verb-patterner) as its dependent, and has direction \emph{opposite} to the auxiliary-verb relation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}. Therefore, this relation is \emph{anti-correlated} with the verb-object relation, while \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}} is \emph{correlated}.  For simplicity, we display this as a corelation in this table.}

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu} we report the results of our two previous, preregistered, simulations\footnote{\url{http://aspredicted.org/blind.php?x=8gp2bt}, \url{https://aspredicted.org/blind.php?x=bg35x7}. For the results of the locality simulations, see the DLM results in Table~\ref{tab:all-predictions-1}, see also Section~\ref{sec:DLM}.} together with results from the main experiment.
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) one simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.

\textcolor{red}{TODO decide on formatting}

@@

\caption{Results from optimization experiments for different values of $\lambda$, including our two previous preregistered experiments. For comparison, we also show results for $\lambda=0$, corresponding to optimizing for parseability only (same results as reported in Tables (\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})). For $\lambda=0.9$, we report results from one preliminary preregistered experiment (left) and the final experiment (right). For $\lambda=1.0$, we report the other preliminary preregistered experiment.
Giving similar weight to parseability and predictability -- that is, $\lambda$ close to $1$ -- results in more accurate predictions than choosing a small value of $\lambda$ such as $\lambda=0.0$. Note that $\lambda$ cannot take values smaller than zero, or greater than one, see Section \ref{sec:lambda}.
}\label{table:corr-resu-previous}
\end{table}

@@

We validated the extracted grammars by comparing the dominant orders of six syntactic relations that are also annotated in the World Atlas of Linguistic Structures (WALS, \cite{haspelmath2005world}).
Among the eight Greenbergian correlations that we were able to test, five are annotated in WALS.
In Table~\ref{tab:grammars-wals}, we compare our grammars with WALS on these five relations, and the verb-object relation.

WALS has some data for all but seven languages.\footnote{Serbian and Croatian are listed as a single language Serbian-Croatian in WALS. In the table, we compare those with the grammar we extracted for Croatian, noting that it fully agrees with the Serbian grammar.}
WALS has data for 74\% of the entries, and lists a dominant order for 91\% of these.
The grammars we extracted from the corpora agree with WALS in 96~\%  of these cases.

@@

Recall the definition of efficiency
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)} 
\end{equation}
\begin{equation}
	R_{Pred} := - H[\mathcal{U}] = \sum_{u} p(u) \log p(u)
\end{equation}
where $t \sim \mathcal{T}$ is a distribution over syntactic structures, and $u \sim p_{\mathcal{L}_\theta}(u|t)$ denotes the corresponding linearized sentences.

@@

Using this, we can rewrite Efficiency~(\ref{eq:efficiency}), for a given grammar $\theta$, equivalently into the parseability and predictability achieved with the best parser and language models:

@@

For gramar optimization, we apply Early Stopping~\citep{prechelt1998early} as follows. We compute a Monte-Carlo estimate of the objective on the UD development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate.
This is a standard technique for optimizing neural models, and helps prevent overfitting to the specific corpora used.

The choice of optimization methods and the stopping criterion were fixed before we investigated word order correlations.

\paragraph{Optimized Grammars}
Recall that, for each language, we created 8 optimized languages for each optimization criterion.
We enforced balanced distribution of object-verb and verb-object ordering among optimized languages by fixing $a_\tau$ for the \textit{obj} dependency to be 0.0 in four of these languages, and 1.0 in the other four.
This maximizes statistical precision in detecting and quantifying correlations between the \textit{obj} relation and other relations.

@@

\textcolor{red}{phantom cite}\citep{diessel2001ordering, dryer1980positional, wals}

>>>>

\textcolor{red}{where do we say that efficiency etc are reported on the held-out set?}

@@

\paragraph{Per-Language Results}
In Figure~\ref{fig:pareto-per-lang}, we show the predictability-parseability planes for every one the 51 languages, together with Pareto frontiers estimated from optimized grammars.
We z-transformed on the level of individual languages, normalizing the mean and SD parseability and predictability of the (1) real grammar, (2) the mean of predictability and parseability of all random grammars, (3) the grammar optimized for efficiency (at $\lambda =0.9$, see Section~\ref{sec:lambda}), (4) grammar optimized for parseability only, and (5) grammar optimized for predictability only.
Within each language, we estimate the Pareto curve based on the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section~\ref{sec:lambda}), (2) parseability, and (3) predictability.
Figure (REF) in the main paper shows the average of these per-language plots.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars. All data are $z$-scored. Note that in a few languages, the real grammar is at a position slightly \emph{beyond} the estimated Pareto frontier. This reflects noise in the optimization process due to use of stochastic gradient descent, i.e., the actual Pareto frontier might be somewhat further away from the baselines than the estimated one.}\label{fig:pareto-per-lang}
\end{figure}

\paragraph{Robustness}
In the main paper, we tested whether real grammars are more efficient than the mean of random grammars, using a $t$-test.
We also conducted the analysis using a Binomial test (one-sided), testing whether the real gramar is more efficient than the \emph{median} of random grammars, avoiding any distributional assumption on the random grammars.
As before, we used Hochberg's step-up procedure (Note that the tests for different languages are independent, as different random grammars are evaluated for each language), with $\alpha=0.05$.
In this analysis, real grammars improved in parseability for 80\% of languages, in predictability for 69\% of languages, and in either of both in 92\% of languages ($p<0.05$, with Bonferroni correction).
In Table~\ref{fig:pareto-per-lang-stats}, we provide per-language results for the $t$-tests and binomial tests.

@@

In Figure~\ref{fig:corr-eff-corr}, we plot efficiency, parseability, and predictability (all are $z$-scored within language, as in Study 1) as a function of the number of satisfied correlations, for the real grammars of the 51 languages.

@@

We show predictions for the eight correlations on the level of individual languages in Figure~\ref{fig:per-lang}.
We obtained these predictions for individual languages and each of the eight relations as follows.
For each language and each of the objective functions (efficiency, predictability, parseability), we considered the optimized grammar that yielded the best value of this objective function among the eight optimized grammars (i.e., the grammar where the optimization procedure had been most successful).
We interpreted this grammar as verb-object or object-verb depending on the order in the real grammar of the language.

@@

Note that the Bayesian analysis  also estimates a posterior distribution of the number of satisfied correlations (see Figure~\ref{fig:posterior}), providing an elegant solution to the multiple-comparisons problem arising from analysing the eight correlations together.

@@

In Figure~\ref{fig:posterior}, we plot the posterior distribution of the number of correlations predicted to hold in most optimized grammars, as obtained from the Bayesian regression.
For each posterior sample, we say that the $j$-th correlation holds if the value of $\alpha_j$ in that posterior sample is positive.
In the figure, we plot the fraction of posterior samples in which a given number of correlations is satisfied.
In addition to grammars optimized for efficiency, we also report the result for grammars optimized for predictability and for parseability alone.
Efficiency predicts all eight correlations with high posterior probability; predictability and parseability alone do not.

\begin{figure}[ht]
	\begin{center}
	\includegraphics[width=0.98\textwidth]{../results/correlations/figures/posterior-satisfied-universals-together-large-three.pdf}
	\end{center}
	\caption{Posterior of the number of correlations correctly predicted by efficiency and its components, in the Bayesian multivariate mixed-effects logistic regression with random effects for languages and language families.}\label{fig:posterior}
\end{figure}

@@

We considered all UD syntactic relations occurring in at least two of the 51 languages.
In Table~\ref{tab:all-predictions-1}, we present the data for the eight correlations discussed in the main paper, and for those other relations for which the typological literature provides data.\footnote{The \emph{aux} syntactic relation in UD has the auxiliary (verb-patterner) as its dependent, and has direction \emph{opposite} to the auxiliary-verb relation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}. Therefore, this relation is \emph{anti-correlated} with the verb-object relation, while \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}} is \emph{correlated}.  For simplicity, we display this as a corelation in this table.}
Additionally, in Table~\ref{tab:all-predictions-2} we present data for the other UD relations, for which either no typological data is available, or which are not linguistically meaningful.

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM} for discussion).
This has been suggested as a property of efficient word order in prior work, and, indeed, we found that DLM is predicted by efficiency optimization (Section~\ref{sec:DLM}).
Results in Table~\ref{tab:all-predictions-1} also show that optimizing for DLM makes predictions similar to efficiency optimization. See Section~\ref{sec:DLM} for further discussion for further discussionn.

@@

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu-previous} we report the results of our two previous, preregistered, simulations\footnote{\url{http://aspredicted.org/blind.php?x=8gp2bt}, \url{https://aspredicted.org/blind.php?x=bg35x7}. For the results of the locality simulations described in the first preregistration, see the DLM results in Table~\ref{tab:all-predictions-1}, see also Section~\ref{sec:DLM} for discussion.} together with results from the main experiment.
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) one simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.

@@

\caption{Results from optimization experiments for different values of $\lambda$, including our two previous preregistered experiments. For comparison, we also show results for $\lambda=0$, corresponding to optimizing for parseability only (same results as reported in Tables (\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})). For $\lambda=0.9$, we report results from one preliminary preregistered experiment (left) and the final experiment (right). For $\lambda=1.0$, we report the other preliminary preregistered experiment.
Giving similar weight to parseability and predictability -- that is, $\lambda$ close to $1$ -- results in more accurate word order predictions than choosing a small value of $\lambda$ such as $\lambda=0.0$. Note that $\lambda$ cannot take values smaller than zero, or greater than one, see Section \ref{sec:lambda}.
}\label{table:corr-resu-previous}
\end{table}

@@

We validated the extracted grammars by comparing the dominant orders of six syntactic relations that are also annotated in the World Atlas of Linguistic Structures (WALS, \cite{haspelmath2005world}).
Among the eight Greenbergian correlations that we were able to test, five are annotated in WALS.
In Table~\ref{tab:grammars-wals}, we compare our grammars with WALS on these five relations, and the verb-object relation.
WALS has data for 74\% of the entries\footnote{Serbian and Croatian are listed as a single language Serbian-Croatian in WALS. In the table, we compare those with the grammar we extracted for Croatian, noting that it fully agrees with the Serbian grammar.}, and lists a dominant order for 91\% of these.
The grammars we extracted from the corpora agree with WALS in 96~\%  of these cases.

@@

Recall the definition of efficiency
\begin{equation}\label{eq:efficiency-recall}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)} 
\end{equation}
\begin{equation}
	R_{Pred} := - H[\mathcal{U}] = \sum_{u} p(u) \log p(u)
\end{equation}
where $t \sim \mathcal{T}$ is a distribution over syntactic structures, and $u \sim p_{\mathcal{L}_\theta}(u|t)$ denotes the corresponding linearized sentences.

@@

Using this, we can rewrite Efficiency~(\ref{eq:efficiency-recall}), for a given grammar $\theta$, equivalently into the parseability and predictability achieved with the best parser and language models:
\begin{equation}\label{eq:efficiency-rewrite}

@@

All corpora have a predefined split in training and held-out (development) sets.
We use the training set for optimizing parameters, and apply Early Stopping~\citep{prechelt1998early} using the held-out set.

For \key{estimating the parseability or predictability} of a given grammar, we optimize the neural model on data ordered according to this grammar, and report the parseability/predictability on the held-out set to avoid overfitting to the training set.
For Early Stopping, we evaluate on the held-out set at the end of every epoch. %run through the training set.

For \key{optimizing grammars}, we jointly apply gradient descent to the grammar parameters and the neural models, using the gradient estimator (\ref{eq:dtheta}-\ref{eq:dpsi}).
For Early Stopping, we evaluate on the held-out set in intervals of 50,000 sentences, using a Monte-Carlo estimate of $R_{\textit{Eff}}^{\theta, \phi, \psi}$ (\ref{eq:efficiency-rewrite}), sampling a single linearized sentence for each syntactic structure in the held-out set.
When reporting the parseability/predictability of an optimized grammar, we evaluate these values for its fully deterministic version (Section~\ref{sec:diff-gramm}) to allow fair comparison with baseline grammars.

%We compute a Monte-Carlo estimate of the objective on the UD development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate.
%This is a standard technique for optimizing neural models, and helps prevent overfitting to the specific corpora used.

The choice of optimization methods and the stopping criterion were fixed before we investigated language efficiency or word order correlations.

\paragraph{Optimized Grammars}
As described in the main paper, for each language, we created 8 optimized languages for each optimization criterion.
We enforced balanced distribution of object-verb and verb-object ordering among optimized languages by fixing $a_\tau$ for the \textit{obj} dependency to be 0.0 in four of these languages, and 1.0 in the other four.
This maximizes statistical precision in detecting and quantifying correlations between the \textit{obj} relation and other relations.

@@

\nocite{diessel2001ordering, dryer1980positional, wals}
